Investigation of Multi-Angle Detection Schemes for Light Scattering of Optically Trapped Asymmetric Colloidal Particles

### 2023-10-25 Latex files/Manuscripts Scattering/Manuscript_Optical Communications.tex
Update to Latex files sub folder

\documentclass[final,  3p]{elsarticle}

@@

\title{Investigation of Limited Detection Schemes for Light Scattering of Optically Trapped Asymmetric Particles}

@@

\cortext[cor1]{Corresponding author}
\affiliation[aff1]{organization={Department of Chemical Engineering,
			University of Strathclyde},
            addressline={75 Montrose Street}, 
            city={Glasgow},
            postcode={G1 1XL}, 
            country={Scotland}}

\begin{flushright}
	Daniel Maciver \\
	daniel.maciver.2016@uni.strath.ac.uk \\
	University of Strathclyde, \\
	75 Montrose Street, \\
	Glasgow, \\
	G1 1XL \\
	07793445441 
\end{flushright}

\begin{abstract}
  % \justifying
  %
  While optical trapping is a well understood method for force
  transduction and detection, further characterisation of trapped entities using light-scattering
  poses a two-fold challenge --- one experimental, concerning the optimal
  arrangement of light detectors to gather data and minimising signal noise, and the other theoretical, involving
  solving of the inverse light scattering problem in order to interpret this data. Experimentally, combining static
  light scattering techniques with optical trapping poses significant
  engineering challenges due to the space constraints in a
  conventional optical trapping setup.  We propose here a plausible
  scenario of detecting scattered light from an optically trapped
  asymmetric microstructure using a novel, multi-angle, optical-fibre
  based detection scheme and demonstrate how a Bayesian inference
  based analysis of the data, combined with a neural-network trained on data simulated to mimic light scattering
  detection signals in such scenarios, may be used for solving the
  inverse light scattering problem and characterising complex trapped
  entities.  To demonstrate the method we discuss its application to
  measuring the instantaneous orientations of a trapped asymmetric microsphere
  dimer. We argue that the method can be extended to
  determine any characteristics of the trapped microstructure that
  influence the light scattering pattern.
\end{abstract}

@@

\section{Introduction}
\label{sec:Intro}

Since their invention in the late 1980s, optical tweezers have found application in experiments ranging from single molecule biophysics \cite{Bustamante2021Biophysics} to testing the fundamental assumptions of quantum mechanics \cite{yin2013large}, thanks mainly to the ability of the tweezer to transduce and detect forces on the order of a few pico-newtons. Going beyond forces, further structural, dynamic and chemical characterisation of complex trapped entities could provide useful information, as demonstrated in areas such as metrology \cite{arita2020coherent} and colloidal aggregation \cite{burns1990optical}.  Spectroscopic techniques such as Raman scattering \cite{gupta2014raman} have been used for chemical characterisation of trapped objects, while dynamical characterisation has been demonstrated using data from tweezer's Quadrant Photo Detector (QPD) by following the centre-of-mass Brownian motion of the trapped entity \cite{friedrich2012tuning} and measuring rotation of the centre-of-mass\cite{yifat2021facile}.  A recent work aimed at characterising trapped entities demonstrated how neural networks maybe trained to distinguish between optically trapped micro-beads of different size and material by means of a principal component analysis of the forward scattered light detected using a QPD \cite{Carvalho_2023}. A more direct, albeit cumbersome attempt at detecting scattered light from trapped biological cells was attempted in \cite{Watson_2023} where the experimental cuvette was placed inside an elliptical mirror that directed light scattered from the trapped biological-cell onto a photodetector via a rotating aperture that helped select the scattering-angle. Thus, past studies on optical trapping have focused on either on particle trapping to study trapping dynamics, or on the characterisation of particles with relatively simple complex trapping dynamics.

While both \cite{Carvalho_2023} \cite{Watson_2023} demonstrate some ability to characterise trapped entities, \cite{Carvalho_2023} is perhaps best suited to characterise micron-sized particles with simple trapping dynamics, and \cite{Watson_2023} talks about an experimental setup that is difficult to adopt and suffers from a low-bandwidth that might not be best suited for monitoring dynamics. A light scattering detection scheme built around an optical trap that is both easy to implement and has the advantage of high-bandwidth was demonstrated by Safran and co-workers in \cite{Bar-Ziv_1998}, where a single-mode optical fibre was aligned to detect the scattered light from a trapped bead and study its Brownian motion - commonly now refereed to as Localised dynamic light scattering (LDLS). This was later expanded upon by \cite{PhysRevE.65.041921} by collecting back scattered light to characterise the Stokes friction coefficient as a function of trapping depth. While both papers provided dynamical information, structural information about the trapped bead was precluded as the scattered wave was only measured across a small angular range. Furthermore, the main drawback to a LDLS is that it struggles to characterise asymmetric particles such as nano rods or dimers; in which case the Brownian motion is no longer simply translational, but rotational about the centre of mass \cite{yin2013large}.

One way of mitigating this is to remove translational fluctuations from the analysis; by monitoring its the scattering, Cang and co-workers were able to spatially fix a gold nanorod at the centre of trapping laser by moving the sample plane accordingly, to an accuracy of 200 nm \cite{Cang2006}. However, this has limited application's to anisotropic scatters such as biological matter where the internal structure of many cells makes them inherently anisotropic \cite{Watson_2023}. Thus even if the Cartesian coordinates are relatively fixed characterisation of the scattering pattern will have to separate the contributions of internal structure and orientation. In this work, we propose a scheme that expands on the technique in \cite{Bar-Ziv_1998} to detect scattered light simultaneously at a number of angles (Figure 1), combined with a novel Bayesian inference-based analysis technique, to enable interpretation of the resulting multi-angle data from an anisotropic, asymmetric scatterer, as well as optimisation to provide maximal information from the signal.

To demonstrate the analysis, we study a simple, illustrative example of an anisotropic scattering entity \emph{i.e.} an asymmetric dimer. As a paradigmatic example of extracting structural/dynamic information from the resulting scattering data, we explore how to estimate the dimer's instantaneous orientation from the scattering signals using Bayesian inference, as well as how to optimise the analysis by implementing 'prior knowledge' to obtain the most reliable estimate. As an example of the relatively sparse literature on measuring orientations of complex trapped objects, \cite{raudsepp2022estimating} employs imaging to study the orientation of trapped dimers: scattering can give more quantitative and potentially more rapid time-resolved information, but only of course if the scattering signals can be interpreted. Here, we first train a neural network to effectively identify the mapping between scattering signals and dimer orientation, by calculating the scattering signal from a simulated asymmetric dimer undergoing Brownian motion in an optical trap and mapping to the known instantaneous orientation of the simulated dimer. We then show how Bayesian inference can be used to optimise extraction of the true dimer orientation from the light scattering signals. Furthermore we demonstrate how the model's performance when dealing with signal noise, a common problem when analysing scattering behaviour.

\begin{center}
		{\bf{Fig 1a, 1b}}
\end{center}
\begin{figure} [b]
\centering
\begin{subfigure}{0.45\textwidth}
	\subcaption{}
	\includegraphics[width=\textwidth, height=0.25\textheight]{./Images/fig1a.png}
\end{subfigure}
\begin{subfigure}{0.45\textwidth}
	\subcaption{}
	\includegraphics[width=\textwidth, height=0.25\textheight]{./Images/fig1b.png}
\end{subfigure}
\caption{\label{fig:setup}
  %
  Proposed experimental set up for scattering measurements from an object in an optical trap. The probe beam for scattering measurements is incident perpendicular to the trapping laser propagation direction. a) Side view. b) Top view. Note that three of the detector fibres are co-planar with the incident probe beam, while the fourth detector is placed out of the plane (see Sec~\ref{sec:detectors}).
%
}
\end{figure}

@@

\begin{center}
	\bf{Fig 2a, 2b}
\end{center}
\begin{figure}[h]
	\centering
	\begin{subfigure}{0.49\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig2a.png}
	\end{subfigure}
	\begin{subfigure}{0.49\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig2b.png}
	\end{subfigure}
	\caption{(a)Example dimer in orientation $\bar{\bf s}$, (b) 30 Reference orientations represented by vectors pointing from [0,0,0] to each point}
	\label{fig:dimer}
\end{figure}

@@

\begin{center}
	\bf{Fig 3a, 3b}
\end{center}
\begin{figure}[h]
	\centering
	\begin{subfigure}{0.4\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig3a.png}
	\end{subfigure}
	\begin{subfigure}{0.4\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig3b.png}
	\end{subfigure}
	\caption{(a) Distribution of orientation vectors and (b) their respective scattering signals. Points are coloured according to their distance from the centre of each cluster (red points centred around [$0.00, 0.00, 1.00$], blue points centred at [$0.71, 0.00, 0.71$])}
	\label{fig:mixing}
\end{figure}

@@

\begin{center}
	\bf{Fig 4a, 4b}
\end{center}
\begin{figure}[h]
	\centering
	\begin{subfigure}{0.49\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig4a.png}
	\end{subfigure}
	\begin{subfigure}{0.4\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig4b.png}
	\end{subfigure}
	\caption{Expected scattered signals from reference orientations -see fig~\ref{fig:dimer} - when: (a) all three detectors are in the X-Y plane, (b) when 1 detector is raised out of the X-Y plane.}
	\label{fig:detectors}
\end{figure}

@@

\begin{center}
	\bf{Fig 5}
\end{center}
\begin{figure}[h]
	\centering
	\includegraphics[width=0.5\textwidth]{./Images/fig5.png}
	\caption{Model's estimation of dimer orientation over the simulation time, assuming uniform prior $p(\hat{\bf n}_\alpha)$, broken up into x, y, and z components for clarity. Blue line denotes the best result we can achieve (the reference orientation $\hat{\bf n}_{best}$ that is closest to the actual orientation), orange line denotes the result provided by eq~\ref{eq:bayes}: where the orange line is not visible, the model's prediction agrees with $\hat{\bf n}_{best}$. Dotted black line is the instantaneous orientation $\hat{\bf s}$.}
	\label{fig:uniform}
\end{figure}

@@

\begin{center}
	\bf{Fig 6}
\end{center}
\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{./Images/fig8a.png}
\caption{\label{fig:biased}
%
Estimation of dimer orientation with $p(\hat{\bf n}_\alpha)$ defined
by Eq~\eqref{eq:boltz}.  Blue line denotes the best result we can
achieve, orange line denotes the result provided by eq~\ref{eq:bayes}.
Dotted black line is the instantaneous orientation $\hat{\bf s}$ (see Section~\ref{sec:Bayes}).
  %
}
\end{figure}

@@

\begin{center}
	\bf{Fig 7a, 7b}
\end{center}
\begin{figure}[h]
	\centering
	\begin{subfigure}{0.45\textwidth}
		\subcaption{}
		\includegraphics[width =\textwidth]{./Images/fig7a.png}
	\end{subfigure}
	\begin{subfigure}{0.45\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig7b.png}
	\end{subfigure}
	\caption{Simulation results of: (a) the dimer's orientation vector with time, (b) the dimer's [x,y,z] position with time.}
	\label{fig:motion}
\end{figure}

@@

\begin{center}
	\bf{Fig 8a, 8b, 8c}
\end{center}
\begin{figure}[h]
	\centering
	\begin{subfigure}{0.33\textwidth}
			\subcaption{}
			\includegraphics[width =\textwidth]{./Images/fig8a.png}
	\end{subfigure}
	\begin{subfigure}{0.31\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig8b.png}
	\end{subfigure}
	\begin{subfigure}{0.31\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig8c.png}
	\end{subfigure}
	\caption{Model estimates of orientation when neural net has been trained on dimer of size ratio: (a) 1:2 $[F(K_l)=9.456]$, (b) 1:2.05 $[F(K_l)=1.324]$, (c) 1:1.95 $[F(K_l)=1.325]$ ($n_{refs} = 30$)}
	\label{fig:size}
\end{figure}
As can be seen from Fig~\ref{fig:size}  even the slightest change in size ratio makes a very significant difference to the performance of our model. This amounts to just over $100 \ nm$ in the dimer's overall size, yet results in our model being correct from over 90 \% of the time to now as low as 30 \%. This highlights the importance of correctly sizing trapped entities before performing any in depth analysis of the scattering pattern, as even the slightest deviation can have a serious impact. We addressed this by increasing the number of available reference orientations from 30 to 126 (following the same procedure as given by \cite{Rey2006} to evenly space out the coordinates) and increasing the weighting factor in Eq~\ref{eq:boltz}. While this didn't have a significant improvement on the overall accuracy of the model, in the worst case having a slight increase from 30.5 \% to 40.3 \%, it did help to significantly reduce the magnitude between our model's estimations and the dimer's motion as seen below in Fig~\ref{fig:refs}.

\begin{center}
	\bf{Fig 9a, 9b, 9c}
\end{center}
\begin{figure}[h]
	\centering
	\begin{subfigure}{0.31\textwidth}
		\subcaption{}
		\includegraphics[width =\textwidth]{./Images/fig9a.png}
	\end{subfigure}
	\begin{subfigure}{0.31\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig9b.png}
	\end{subfigure}
	\begin{subfigure}{0.31\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig9c.png}
	\end{subfigure}
	\caption{Model estimates of orientation when neural net has been trained on dimer of size ratio: (a) 1:2 $[F(K_l)=11.756]$, (b) 1:2.05 $[F(K_l)=1.233]$, (c) 1:1.95 $[F(K_l)=2.128]$, ($n_{refs} = 126$)}
	\label{fig:refs}
\end{figure}

@@

\begin{center}
	\bf{Fig 10a, 10b, 10c}
\end{center}
\begin{figure}[h]
	\centering
	\begin{subfigure}{0.32\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig10a.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig10b.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig10c.png}
	\end{subfigure}
	\caption{Model prediction for signal error of (a) $1\%$ ~  $[F(K_l)=7.246]$, (b) $15\%$ $[F(K_l)=0.511]$, and (c) $25\%$ ~  $[F(K_l)=0.536]$.}
	\label{fig:epsilon}
\end{figure}

@@

\begin{center}
	\bf{Fig 11a, 11b, 11c}
\end{center}
\begin{figure}[h]
	\centering
	\begin{subfigure}{0.32\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig11a.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig11b.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig11c.png}
	\end{subfigure}
	\caption{Model prediction for signal error of (a) $1\%$ $[F(K_l)=4.823]$,  (b) $15\%$ $[F(K_l)=1.494]$, and (c) $25\%$ $[F(K_l)=0.882]$, time averaged over $1\,{\rm ms}$}
	\label{fig:time average}
\end{figure}
This resulted in a reduction in the overall signal noise and provided a higher degree of accuracy for our model. There appears to be no clear correlation between the length over which we time average and the performance of our model. Time averaging over every 0.05s resulted in a drastically worse performance; this is due to the fact that over longer time periods there is greater uncertainty regarding how the dimer's orientation has changed, thus tracking the instantaneous orientation becomes harder for the neural network. Fortunately, time averaging even over $1\,{\rm ms}$ seems to provide a satisfactory estimation of the dimer's angular dynamics within the optical trap.

@@

We have developed a method for measuring the dynamics of an optically-trapped 'complex object' based purely on limited measurements of the object's light scattering at a small number of detection angles. We demonstrate the method using the orientation of an asymmetric dimer as the dynamical variable and object of interest, but in principle the model can be applied to any characteristic that impacts the light scattering pattern produced by a trapped entity. The MSTM package is a flexible tool for calculating the light scattering of complex objects using a representation of the object as a set of micro-particles, enabling training of a neural network to enable categorisation of the mapping between scattering and trapped object characteristics. By taking account of knowledge about the physically realistic behaviour of the trapped object and the characteristics of the trap (which impact the dynamics of the object), the Bayesian inference method can be refined to provide a reliable estimation of object characteristics of interest, even in the presence of measurement noise. Fundamentally the inverse scattering problem is difficult to solve since the mapping between object characteristics and scattering can be highly complex: but Bayesian inference based on neural network estimation of the mapping provides a powerful method for practical applications, extending the use of optical trapping beyond measuring microscopic force response toward detailed structural and dynamic information about complex trapped entities.\\

>>>>

\documentclass[preprint,  3p]{elsarticle}

@@

\title{Investigation of Multi-Angle Detection Schemes for Light Scattering of Optically Trapped Asymmetric Colloidal Particles}

@@

\cortext[cor1]{Corresponding author}
\affiliation[aff1]{organization={Department of Chemical Engineering,
			University of Strathclyde},
            addressline={75 Montrose Street}, 
            city={Glasgow},
            postcode={G1 1XL}, 
            country={Scotland}}
\begin{abstract}
  % \justifying
  %
  Optical trapping is a well understood method for transduction and detection
  of forces on trapped colloidal particles, these trapped entities can be
  further characterised using light-scattering, posing a two-fold challenge:
  one experimental, concerning the optimal arrangement of detectors to gather
  data and minimise signal noise, and the other theoretical, involving solving
  of the inverse scattering problem in order to interpret light scattering
  data to determine size, shape, or orientation of the trapped object.
  Experimentally, combining static light scattering techniques with optical
  trapping poses significant engineering challenges due to the space
  constraints in a conventional optical trapping setup. We investigated a
  plausible scenario of detecting scattered light from an optically trapped
  asymmetric particles using a novel, multi-angle, optical-fibre based
  detection scheme, we show how a Bayesian inference-based analysis of the
  data, combined with a neural-network trained on data simulated to mimic
  light scattering detection signals in such scenarios, can be used for
  solving the inverse light scattering problem and characterising colloidal
  trapped entities. To demonstrate the method, we discuss its application to
  measuring the instantaneous orientations of a trapped asymmetric microsphere
  dimer and determine the minimum number of detectors required for a reliable
  estimation in the presence of signal noise. This approach can be extended to
  determine any characteristics of the trapped microstructure that influence
  the light scattering pattern, including size and shape of colloidal objects. 
\end{abstract}

%\begin{graphicalabstract}
%\begin{figure} [h]
%	\centering
%	\begin{subfigure}{0.45\textwidth}
%		\includegraphics[width=\textwidth, height=0.25\textheight]{./Images/fig1a.png}
%	\end{subfigure}
%	\begin{subfigure}{0.45\textwidth}
%		\includegraphics[width=\textwidth, height=0.25\textheight]{./Images/fig1b.png}
%	\end{subfigure}
%\end{figure}
%\end{graphicalabstract}

@@

\section{Introduction}
\label{sec:Intro}
Since their invention in the late 1980s, optical tweezers have found application in experiments ranging from single molecule biophysics \cite{Bustamante2021Biophysics} to testing the fundamental assumptions of quantum mechanics \cite{yin2013large}, thanks to the Brownian dynamics of colloidal systems to the tweezer can transduce and detect forces down to the order of a few pico-newtons. Going beyond forces, further structural, dynamic and chemical characterisation of complex trapped entities could provide useful information, as demonstrated in areas such as metrology \cite{arita2020coherent} and colloidal science \cite{burns1990optical}. Spectroscopic techniques such as Raman scattering \cite{gupta2014raman} have been used for chemical characterisation of trapped objects, while dynamical characterisation has been demonstrated using data from tweezer's Quadrant Photo Detector (QPD) by following the centre-of-mass Brownian motion of the trapped entity \cite{friedrich2012tuning} and measuring rotation of the centre-of-mass \cite{yifat2021facile}. A recent work aimed at characterising trapped entities demonstrated how neural networks can be trained to distinguish between optically trapped micro-beads of different size and material by means of a principal component analysis of the forward scattered light detected using a QPD \cite{Carvalho_2023}. A more direct, albeit cumbersome attempt at detecting scattered light from trapped biological cells also was attempted \cite{Watson_2023} where the experimental cuvette was placed inside an elliptical mirror that directed light scattered from the trapped biological-cell onto a photodetector via a rotating aperture that helped select the scattering-angle. Thus, past studies on optical trapping have focused on either on particle trapping to study trapping dynamics, or on the characterisation of particles with relatively simple complex trapping dynamics.

While both \cite{Carvalho_2023} and \cite{Watson_2023} demonstrate some ability to characterise trapped entities, \cite{Carvalho_2023} is perhaps best suited to characterise micron-sized particles with simple trapping dynamics, and \cite{Watson_2023} describes an experimental setup that is difficult to adopt and suffers from a low bandwidth that might not be best suited for monitoring dynamics. A light scattering detection scheme built around an optical trap that is easier to implement and has the advantage of high bandwidth was demonstrated by Safran and co-workers in \cite{Bar-Ziv_1998}, where a single-mode optical fibre was aligned to detect the scattered light from a trapped bead and study its Brownian motion, commonly now referred to as Localised Dynamic Light Scattering (LDLS). This was later expanded upon \cite{Viana2002} by collecting back scattered light to characterise the Stokes friction coefficient as a function of trapping depth. While both papers provided dynamical information, structural information about the trapped bead was not available as the scattered light was only measured across a small angular range. Furthermore, the main drawback to a LDLS is that it cannot be used to characterise asymmetric particles such as dimers or more complex aggregates; in which case the Brownian motion is no longer simply translational, but rotational about the centre of mass \cite{yin2013large}.

One way of mitigating this is to remove translational fluctuations from the analysis; by monitoring it’s scattering pattern, Cang and co-workers were able to spatially fix a gold nanorod at the centre of trapping laser by moving the sample plane accordingly, to an accuracy of 200 nm \cite{Hu2006}. However, this has limited applications to anisotropic scatterers such as biological matter where the internal structure of many cells makes them inherently anisotropic \cite{Watson_2023}. Thus, even if the Cartesian coordinates are fixed, characterisation of the scattering pattern will have to separate the contributions of size, structure, and orientation.  As an example of the relatively sparse literature on measuring orientations of complex trapped objects, \cite{raudsepp2022estimating} employs imaging to study the orientation of trapped dimers: scattering can give more quantitative and potentially more rapid time-resolved information, but only of course if the scattering signals can be interpreted. In this work, we propose a novel approach that expands on the previous approaches \cite{Bar-Ziv_1998} \cite{Viana2002} to detect scattered light simultaneously at multiple angles (Figure 1), combined with a novel Bayesian inference-based analysis technique, to enable interpretation of the resulting multi-angle data from an anisotropic, asymmetric scatterer, and optimisation to provide maximal information from the detector signals.

To demonstrate this approach, we study an anisotropic scattering entity i.e. an asymmetric dimer, to determine dynamic and structural information about the trapped entity. As a paradigmatic example of extracting information from the scatterer’s scattering data, we explore how to estimate the dimer's instantaneous orientation from the scattering signals using Bayesian inference, as well as how to optimise the analysis by implementing 'prior knowledge' to obtain the most reliable estimate We, first train a neural network to effectively identify the mapping between scattering signals and dimer orientation, by calculating the scattering signal from a simulated asymmetric dimer undergoing Brownian motion in an optical trap and mapping to the 2 known instantaneous orientation of the simulated dimer. We then show how Bayesian inference can be used to optimise our estimation of the true dimer orientation from the light scattering signals. Furthermore, we demonstrate how the model's performance when dealing with signal noise, a common problem when analysing scattering behaviour. This approach can be extended to determine any characteristic that influences the light scattering pattern of trapped colloidal objects including: size, shape, and orientation.

\begin{figure} [h]
\centering
\begin{subfigure}{0.45\textwidth}
	\subcaption{}
	\includegraphics[width=\textwidth, height=0.25\textheight]{./Images/fig1a.png}
\end{subfigure}
\begin{subfigure}{0.45\textwidth}
	\subcaption{}
	\includegraphics[width=\textwidth, height=0.25\textheight]{./Images/fig1b.png}
\end{subfigure}
\caption{\label{fig:setup}
  %
  Proposed experimental set up for scattering measurements from an object in an optical trap. The probe beam for scattering measurements is incident perpendicular to the trapping laser propagation direction. a) Side view. b) Top view. Note that three of the detector fibres are co-planar with the incident probe beam, while the fourth detector is placed out of the plane (see Sec~\ref{sec:detectors}).
%
}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\begin{subfigure}{0.49\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig2a.png}
	\end{subfigure}
	\begin{subfigure}{0.49\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig2b.png}
	\end{subfigure}
	\caption{(a)Example dimer in orientation $\bar{\bf s}$, (b) 30 Reference orientations represented by vectors pointing from [0,0,0] to each point}
	\label{fig:dimer}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\begin{subfigure}{0.4\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig3a.png}
	\end{subfigure}
	\begin{subfigure}{0.4\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig3b.png}
	\end{subfigure}
	\caption{(a) Distribution of orientation vectors and (b) their respective scattering signals. Points are coloured according to their distance from the centre of each cluster (red points centred around [$0.00, 0.00, 1.00$], blue points centred at [$0.71, 0.00, 0.71$])}
	\label{fig:mixing}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\begin{subfigure}{0.49\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig4a.png}
	\end{subfigure}
	\begin{subfigure}{0.4\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig4b.png}
	\end{subfigure}
	\caption{Expected scattered signals from reference orientations -see fig~\ref{fig:dimer} - when: (a) all three detectors are in the X-Y plane, (b) when 1 detector is raised out of the X-Y plane.}
	\label{fig:detectors}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=0.5\textwidth]{./Images/fig5.png}
	\caption{Model's estimation of dimer orientation over the simulation time, assuming uniform prior $p(\hat{\bf n}_\alpha)$, broken up into x, y, and z components for clarity. Blue line denotes the best result we can achieve (the reference orientation $\hat{\bf n}_{best}$ that is closest to the actual orientation), orange line denotes the result provided by eq~\ref{eq:bayes}: where the orange line is not visible, the model's prediction agrees with $\hat{\bf n}_{best}$. Dotted black line is the instantaneous orientation $\hat{\bf s}$.}
	\label{fig:uniform}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{./Images/fig8a.png}
\caption{\label{fig:biased}
%
Estimation of dimer orientation with $p(\hat{\bf n}_\alpha)$ defined
by Eq~\eqref{eq:boltz}.  Blue line denotes the best result we can
achieve, orange line denotes the result provided by eq~\ref{eq:bayes}.
Dotted black line is the instantaneous orientation $\hat{\bf s}$ (see Section~\ref{sec:Bayes}).
  %
}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\begin{subfigure}{0.45\textwidth}
		\subcaption{}
		\includegraphics[width =\textwidth]{./Images/fig7a.png}
	\end{subfigure}
	\begin{subfigure}{0.45\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig7b.png}
	\end{subfigure}
	\caption{Simulation results of: (a) the dimer's orientation vector with time, (b) the dimer's [x,y,z] position with time.}
	\label{fig:motion}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\begin{subfigure}{0.33\textwidth}
			\subcaption{}
			\includegraphics[width =\textwidth]{./Images/fig8a.png}
	\end{subfigure}
	\begin{subfigure}{0.31\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig8b.png}
	\end{subfigure}
	\begin{subfigure}{0.31\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig8c.png}
	\end{subfigure}
	\caption{Model estimates of orientation when neural net has been trained on dimer of size ratio: (a) 1:2 $[F(K_l)=9.456]$, (b) 1:2.05 $[F(K_l)=1.324]$, (c) 1:1.95 $[F(K_l)=1.325]$ ($n_{refs} = 30$)}
	\label{fig:size}
\end{figure}
As can be seen from Fig~\ref{fig:size}  even the slightest change in size ratio makes a very significant difference to the performance of our model. This amounts to just over $100 \ nm$ in the dimer's overall size, yet results in our model being correct from over 90 \% of the time to now as low as 30 \%. This highlights the importance of correctly sizing trapped entities before performing any in depth analysis of the scattering pattern, as even the slightest deviation can have a serious impact. We addressed this by increasing the number of available reference orientations from 30 to 126 (following the same procedure as given by \cite{Rey2006} to evenly space out the coordinates) and increasing the weighting factor in Eq~\ref{eq:boltz}. While this didn't have a significant improvement on the overall accuracy of the model, in the worst case having a slight increase from 30.5 \% to 40.3 \%, it did help to significantly reduce the magnitude between our model's estimations and the dimer's motion as seen below in Fig~\ref{fig:refs}.

\begin{figure}[h]
	\centering
	\begin{subfigure}{0.31\textwidth}
		\subcaption{}
		\includegraphics[width =\textwidth]{./Images/fig9a.png}
	\end{subfigure}
	\begin{subfigure}{0.31\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig9b.png}
	\end{subfigure}
	\begin{subfigure}{0.31\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig9c.png}
	\end{subfigure}
	\caption{Model estimates of orientation when neural net has been trained on dimer of size ratio: (a) 1:2 $[F(K_l)=11.756]$, (b) 1:2.05 $[F(K_l)=1.233]$, (c) 1:1.95 $[F(K_l)=2.128]$, ($n_{refs} = 126$)}
	\label{fig:refs}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\begin{subfigure}{0.32\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig10a.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig10b.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig10c.png}
	\end{subfigure}
	\caption{Model prediction for signal error of (a) $1\%$ ~  $[F(K_l)=7.246]$, (b) $15\%$ $[F(K_l)=0.511]$, and (c) $25\%$ ~  $[F(K_l)=0.536]$.}
	\label{fig:epsilon}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\begin{subfigure}{0.32\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig11a.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig11b.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\subcaption{}
		\includegraphics[width=\textwidth]{./Images/fig11c.png}
	\end{subfigure}
	\caption{Model prediction for signal error of (a) $1\%$ $[F(K_l)=4.823]$,  (b) $15\%$ $[F(K_l)=1.494]$, and (c) $25\%$ $[F(K_l)=0.882]$, time averaged over $1\,{\rm ms}$}
	\label{fig:time average}
\end{figure}
This resulted in a reduction in the overall signal noise and provided a higher degree of accuracy for our model. There appears to be no clear correlation between the length over which we time average and the performance of our model. Time averaging over every 0.05s resulted in a drastically worse performance; this is due to the fact that over longer time periods there is greater uncertainty regarding how the dimer's orientation has changed, thus tracking the instantaneous orientation becomes harder for the neural network. Fortunately, time averaging even over $1\,{\rm ms}$ seems to provide a satisfactory estimation of the dimer's angular dynamics within the optical trap.

From the above discussion it’s clear that estimation of the dimer’s orientation is a problem that can be endlessly tuned to fully maximise our end result. Here we simplify the problem somewhat by employing a relatively small finite number of 'reference orientations' to map between scattering and dimer orientation: the precision of estimation could be improved by utilising a greater number of reference orientations, although there remains a balance between the realisable precision of orientation estimate and the noise level of the scattering measurement. Another avenue to further explore would be using the method to optimise the choice of detection angles, essentially to find the region in the mapping between measured scattering and orientation that others the best degree of confidence through optimal separation of scattering signals for distinct orientations. For sequences of data such as dynamic measurements, a further potential enhancement would be to consider more complex correlations based on prior expectations of the dynamics. Here already we improve the method using a non-uniform prior based on only the immediately previous measurement in time (see Section 2.1): considering a non-uniform grouping of reference orientations might result in a better estimation, if we have information regarding the dimer’s preferred axis of rotation.

@@

We have developed a method for measuring the dynamics of an optically-trapped colloidal objects based purely on measurements of the object's light scattering at a small number of detection angles. We demonstrate the method using the orientation of an asymmetric dimer as the dynamic variable and object of interest respectively, but in principle the model can be applied to any characteristic that impacts the light scattering pattern produced by a trapped entity such as size and shape. The MSTM package is a flexible tool for calculating the light scattering of complex objects using a representation of the object as a set of micro-particles, enabling training of a neural network to enable categorisation of the mapping between scattering and trapped object characteristics. By taking account of the physically realistic behaviour of the trapped object and the characteristics of the trap (which impact the dynamics of the object), the Bayesian inference method can be re ned to provide a reliable estimation of object characteristics of interest, even in the presence of measurement noise. Fundamentally, the inverse scattering problem is difficult to solve, since the mapping between object characteristics and scattering can be highly complex. We determined the minimum number of detectors required for a reliable estimation in the presence of measurement noise; furthermore, we demonstrated that the arrangement of these detectors is critical for a reliable estimation of an objects orientation. However, Bayesian inference based on neural network estimation of the mapping provides a powerful method for practical applications, extending the use of optical trapping beyond measuring microscopic force response toward detailed structural and dynamic information about complex trapped entities.
